Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-11-25 manuscript.qmd
fix references

In treating head and neck squamous cell carcinomas (HNSCC) with radiotherapy or surgery, both the primary tumor and clinically detected lymph node metastases are targeted. Current guidelines also include large portions of the neck in the elective clinical target volume (CTV-N) [@gregoire_ct-based_2003;@gregoire_delineation_2014;@gregoire_delineation_2018;@eisbruch_intensity-modulated_2002;@biau_selection_2019;@chao_determination_2002;@vorwerk_guidelines_2011;@ferlito_elective_2009] to mitigate the risk of regional recurrences from untreated microscopic disease undetectable by in-vivo imaging modalities such as computed tomography (CT), magnetic resonance imaging (MRI), or positron emission tomography (PET). However, this approach must balance minimizing the risk of occult disease in the lymphatic drainage region against the toxicity of unnecessarily irradiating healthy tissue.

@@

To develop models for lymphatic tumor progression for all relevant LNLs, including contralateral regions, we compiled a detailed dataset of {{< var data.num_patients >}} patients with newly diagnosed oropharyngeal squamous cell carcinomas (OPSCC) [@ludwig_dataset_2022;@ludwig_multi-centric_2023]. The dataset includes lymph node involvement per LNL for each patient in tabular form, along with primary tumor and patient characteristics such as T-category, subsite, primary tumor lateralization, and HPV p16 status. Patient records were collected from four institutions, and an overview of patient characteristics is provided in @tbl-data-overview.

@@

Pathological involvement is available only for surgically treated patients and for the levels that were dissected. For non-surgical patients, involvement status is determined clinically, i.e. using imaging. To ensure uniformity in analysis, diagnostic information was synthesized into a consensus decision for each patient and LNL. This consensus reflects the most likely state of involvement and accounts for the sensitivity and specificity of various diagnostic modalities, as reported in the literature [@de_bondt_detection_2007;@kyzas_18f-fluorodeoxyglucose_2008].

@@

The complete dataset, including additional patients with tumors in primary sites other than the oropharynx, is publicly accessible. It can be [downloaded from LyProX](https://lyprox.org/patients/dataset), where it is also available for interactive exploration, [from GitHub](https://github.com/rmnldwg/lydata), or [from Zenodo](https://zenodo.org/search?q=lydata). Additional details about the datasets and data format are provided in designated Data-in-Brief publications [@ludwig_dataset_2022;@ludwig_multi-centric_2023]. These publications do not include the latest data from HVH, which will be described in a future publication.

@@

It is possible to provide our model with multiple different diagnostic modalities, each being characterized by different pairs of sensitivity and specificity. However, we instead chose to combine them into a single "consensus" diagnosis before parameter inference, as introduced in @sec-data-consensus. We opted for this because the literature values of sensitivity and specificity [@de_bondt_detection_2007;@kyzas_18f-fluorodeoxyglucose_2008] of imaging modalities like MRI and CT do not plausibly match some of our observations: In the USZ cohort, 78% of OPSCC patients where diagnosed with ipsilateral LNL II involvement via diagnostic imaging. This is virtually impossible with sensitivities around 80% and specificities lower than 100%.

@@

For parameter inference, we used the Python package [`emcee`] [@foreman-mackey_emcee_2013]. It implements efficient MCMC sampling algorithms that employ multiple parallel samplers for affine invariance and better performance on multi-core CPUs. The sample proposal algirothms used by us are based on differential evolution moves [@ter_braak_differential_2008;@nelson_run_2013]. The [`emcee`] library was provided with the likelihood implemented by our [`lymph-model`] Python package.

@@

For a clinical application of the model we are interested in the risk of occult metastases rather than the observed patterns of lymph node involvement. We want to estimate the risk for occult disease in clinically negative LNLs, given the patient's individual diagnosis. In terms of our model, the diagnosis consists of the T-category, the lateralization of the tumor (if it extend over the mid-sagittal plane), and which LNLs are clinically involved based on imaging and possibly fine needle aspiration (FNA). For the results shown in this section, we assume that involvement of a LNL is clinically diagnosed through imaging with a sensitivity of 81% and a specificity of 76% [@de_bondt_detection_2007]. For FNA we assume a specificity of 98% and a sensitivity of 80%. This amounts to the assumption that lymph node involvement confirmed by FNA is almost certainly true involvement, i.e. there are almost no false positive diagnoses.

@@

<!-- @Fig-model-risks shows this predicted risk of occult disease in the contralateral LNLs II, III, and IV for interesting combinations of these three risk factors. This risk is computed *given* a CT diagnosis assessing the per-LNL clinical involvement for which we assumed a sensitivity of 81% and a specificity of 76%. For the involvement in the LNls III and IV contralaterally, we also cosidered how the risk would be affected if the upstream LNL was confirmed to be metastatic using fine needle aspiration (FNA) with a specificity of 98% and a sensitivity of 80% [@de_bondt_detection_2007]. In the figure legends, this is noted e.g. for the orange historgam over the risk of contralateral LNL IV involvement as `c:III FNA+`, meaning that the contralateral LNL III was pathologically confirmed to harbor metastases via FNA.-->

@@

Within the cohort of patients with oropharyngeal tumors, subgroups can be identified whose tumors are located in different subsites, e.g. at the base of the tongue or on the tonsils. The lymphatic metastatic spread patterns of these subgroups may differ slightly. If so, the model could benefit from more fine-grained information about the precise tumor subsite for its training and prediction. We have conducted preliminary work to show that a mixture model may be a promising approach to capture the subsite specific spread patterns [@noauthor_proceedings_2024, @zora231470]. The additionally represents an approach to extend the model towards tumor locations in the oral cavity, hypopharynx, and larynx.

@@

: Specificity and sensitivity values from the literature [@de_bondt_detection_2007;@kyzas_18f-fluorodeoxyglucose_2008]. {#tbl-spec-sens}

>>>>

In treating head and neck squamous cell carcinomas (HNSCC) with radiotherapy or surgery, both the primary tumor and clinically detected lymph node metastases are targeted. Current guidelines also include large portions of the neck in the elective clinical target volume (CTV-N) [@gregoire_ctbased_2003;@gregoire_delineation_2014;@gregoire_delineation_2018;@eisbruch_intensitymodulated_2002;@biau_selection_2019;@chao_determination_2002;@vorwerk_guidelines_2011;@ferlito_elective_2009] to mitigate the risk of regional recurrences from untreated microscopic disease undetectable by in-vivo imaging modalities such as computed tomography (CT), magnetic resonance imaging (MRI), or positron emission tomography (PET). However, this approach must balance minimizing the risk of occult disease in the lymphatic drainage region against the toxicity of unnecessarily irradiating healthy tissue.

@@

To develop models for lymphatic tumor progression for all relevant LNLs, including contralateral regions, we compiled a detailed dataset of {{< var data.num_patients >}} patients with newly diagnosed oropharyngeal squamous cell carcinomas (OPSCC) [@ludwig_dataset_2022;@ludwig_multicentric_2024]. The dataset includes lymph node involvement per LNL for each patient in tabular form, along with primary tumor and patient characteristics such as T-category, subsite, primary tumor lateralization, and HPV p16 status. Patient records were collected from four institutions, and an overview of patient characteristics is provided in @tbl-data-overview.

@@

Pathological involvement is available only for surgically treated patients and for the levels that were dissected. For non-surgical patients, involvement status is determined clinically, i.e. using imaging. To ensure uniformity in analysis, diagnostic information was synthesized into a consensus decision for each patient and LNL. This consensus reflects the most likely state of involvement and accounts for the sensitivity and specificity of various diagnostic modalities, as reported in the literature [@debondt_detection_2007;@kyzas_18ffluorodeoxyglucose_2008].

@@

The complete dataset, including additional patients with tumors in primary sites other than the oropharynx, is publicly accessible. It can be [downloaded from LyProX](https://lyprox.org/patients/dataset), where it is also available for interactive exploration, [from GitHub](https://github.com/rmnldwg/lydata), or [from Zenodo](https://zenodo.org/search?q=lydata). Additional details about the datasets and data format are provided in designated Data-in-Brief publications [@ludwig_dataset_2022;@ludwig_multicentric_2024]. These publications do not include the latest data from HVH, which will be described in a future publication.

@@

It is possible to provide our model with multiple different diagnostic modalities, each being characterized by different pairs of sensitivity and specificity. However, we instead chose to combine them into a single "consensus" diagnosis before parameter inference, as introduced in @sec-data-consensus. We opted for this because the literature values of sensitivity and specificity [@debondt_detection_2007;@kyzas_18ffluorodeoxyglucose_2008] of imaging modalities like MRI and CT do not plausibly match some of our observations: In the USZ cohort, 78% of OPSCC patients where diagnosed with ipsilateral LNL II involvement via diagnostic imaging. This is virtually impossible with sensitivities around 80% and specificities lower than 100%.

@@

For parameter inference, we used the Python package [`emcee`] [@foreman-mackey_emcee_2013]. It implements efficient MCMC sampling algorithms that employ multiple parallel samplers for affine invariance and better performance on multi-core CPUs. The sample proposal algirothms used by us are based on differential evolution moves [@terbraak_differential_2008;@nelson_run_2013]. The [`emcee`] library was provided with the likelihood implemented by our [`lymph-model`] Python package.

@@

For a clinical application of the model we are interested in the risk of occult metastases rather than the observed patterns of lymph node involvement. We want to estimate the risk for occult disease in clinically negative LNLs, given the patient's individual diagnosis. In terms of our model, the diagnosis consists of the T-category, the lateralization of the tumor (if it extend over the mid-sagittal plane), and which LNLs are clinically involved based on imaging and possibly fine needle aspiration (FNA). For the results shown in this section, we assume that involvement of a LNL is clinically diagnosed through imaging with a sensitivity of 81% and a specificity of 76% [@debondt_detection_2007]. For FNA we assume a specificity of 98% and a sensitivity of 80%. This amounts to the assumption that lymph node involvement confirmed by FNA is almost certainly true involvement, i.e. there are almost no false positive diagnoses.

@@

<!-- @Fig-model-risks shows this predicted risk of occult disease in the contralateral LNLs II, III, and IV for interesting combinations of these three risk factors. This risk is computed *given* a CT diagnosis assessing the per-LNL clinical involvement for which we assumed a sensitivity of 81% and a specificity of 76%. For the involvement in the LNls III and IV contralaterally, we also cosidered how the risk would be affected if the upstream LNL was confirmed to be metastatic using fine needle aspiration (FNA) with a specificity of 98% and a sensitivity of 80% [@debondt_detection_2007]. In the figure legends, this is noted e.g. for the orange historgam over the risk of contralateral LNL IV involvement as `c:III FNA+`, meaning that the contralateral LNL III was pathologically confirmed to harbor metastases via FNA.-->

@@

Within the cohort of patients with oropharyngeal tumors, subgroups can be identified whose tumors are located in different subsites, e.g. at the base of the tongue or on the tonsils. The lymphatic metastatic spread patterns of these subgroups may differ slightly. If so, the model could benefit from more fine-grained information about the precise tumor subsite for its training and prediction. We have conducted preliminary work to show that a mixture model may be a promising approach to capture the subsite specific spread patterns [@_proceedings_2024, @zora231470]. The additionally represents an approach to extend the model towards tumor locations in the oral cavity, hypopharynx, and larynx.

@@

: Specificity and sensitivity values from the literature [@debondt_detection_2007;@kyzas_18ffluorodeoxyglucose_2008]. {#tbl-spec-sens}
